Component response rate variation underlies the stability of complex systems

### 2020-03-02 notebook/ms.Rmd
Rearrange some of the text to improve the flow between analytical and numerical simulation sections of the Results

Randomly assembled complex systems can be represented as large square matrices ($\mathbf{M}$) with $S$ components (e.g., networks of species [@Allesina2012] or banks [@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation ($d$, e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma_{M}^{2}$; diagonal elements are set to $d = -1$ [@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $\mathbf{M}$ are negative ($\max\left(\Re(\lambda)\right) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma_{M}\sqrt{SC$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S_{M}$, $C_{M}$, and $\sigma_{M}^{2}$ increase.

@@

Equation \ref{defM} thereby serves as a null model to investigate how variation in component response rate ($\sigma_{\gamma}$) affects complex systems. In the absence of such variation ($\sigma_{\gamma} = 0$), $\gamma$ is set to the identity matrix (diagonal elements all equal 1) and $\textbf{M} = \textbf{A}$. For this baseline condition, $\max(\Re(\lambda))$ can be estimated from five values [@Tang2014b]. These values include (1) system size ($S$), (2) mean self-regulation of components ($d$), (3) mean interaction strength between components ($\mu$), (4) the standard deviation between component interaction strengths ($\sigma_{M}$), and (5) the correlation of interaction strengths between components, $M_{ij}$ and $M_{ji}$ ($\rho$).

When investigating the effect of varying component response rate $\sigma_{\gamma}$ on stability by defining $\textbf{M} = \gamma\textbf{A}$, $S$ remains unchanged. Further, values of $\gamma_{i}$ were sampled such that $E[d]$ and $E[\mu]$ also remained unchanged (in practice, diagonal elements of $\textbf{M}$ were standardised so that mean values were identical before and after adding $\gamma$). What $\sigma_{\gamma}$ does change is the variation in component interaction strengths, and $\rho$. Introducing variation in $\gamma$ increases the total variation in the system, making it more likely that $\textbf{M} = \gamma\textbf{A}$ is unstable. Variation of the off-diagonal elements in $\textbf{M}$ is described by the joint variation of two random variables (to simplify the notation, $\sigma^{2}_{M}$ and $\sigma^{2}_{A}$ refer to variances of off-diagonal elements only),

@@

Note that in my simulations $E[\gamma_{i}] = 1$ and $E[A_{ij}] = 0$, so equation \ref{var_full} can be simplified,

@@

>>>>

Randomly assembled complex systems can be represented as large square matrices ($\mathbf{M}$) with $S$ components (e.g., networks of species [@Allesina2012] or banks [@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation ($d$, e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma_{M}^{2}$; diagonal elements are set to $d = -1$ [@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $\mathbf{M}$ are negative ($\max\left(\Re(\lambda)\right) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma_{M}\sqrt{SC}$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S_{M}$, $C_{M}$, and $\sigma_{M}^{2}$ increase.

@@

Equation \ref{defM} thereby serves as a null model to investigate how variation in component response rate ($\sigma_{\gamma}$) affects complex systems. In the absence of such variation ($\sigma_{\gamma} = 0$), $\gamma$ is set to the identity matrix (diagonal elements all equal 1) and $\textbf{M} = \textbf{A}$.

**Analytical considerations**. The value of $\max(\Re(\lambda))$, and therefore system stability, can be estimated from five properties of $\textbf{M}$ [@Tang2014b]. These properties include (1) system size ($S$), (2) mean self-regulation of components ($d$), (3) mean interaction strength between components ($\mu$), (4) the standard deviation between component interaction strengths ($\sigma_{M}$), and (5) the correlation of interaction strengths between components, $M_{ij}$ and $M_{ji}$ ($\rho$). Positive $\sigma_{\gamma}$ does not change $S$, nor does it necessarily change $E[d]$ or $E[\mu]$. What $\sigma_{\gamma}$ does change is the total variation in component interaction strengths ($\sigma_{M}$), and $\rho$. Introducing variation in $\gamma$ increases the total variation in the system. Variation of the off-diagonal elements in $\textbf{M}$ is described by the joint variation of two random variables,

@@

Given $E[\gamma_{i}] = 1$ and $E[A_{ij}] = 0$, equation \ref{var_full} can be simplified,

@@

The increase caused by $\sigma^{2}_\gamma$ can be visualised from the eigenvalue spectra of $\textbf{A}$ versus $\textbf{M} = \gamma\textbf{A}$ (Figure 1). Given $d = 0$ and $C = 1$, the distribution of eigenvalues of $\textbf{A}$ and $\textbf{M}$ lie within a circle of a radius $\sigma^{2}_{A}\sqrt{S}$ and $\sigma^{2}_{M}\sqrt{S}$, respectively (Figure 1a vs. 1b). When components are self-regulating (e.g., $d = -1$), positive $\sigma^{2}_\gamma$ changes the distribution of eigenvalues, potentially affecting stability (Figure 1c vs. 1d).

Given $\sigma^{2}_\gamma = 0$, $\max(\Re(\lambda))$ decrease linearly with $\rho$ such that,

\begin{equation} \label{rho_stab}
\max(\Re(\lambda)) \approx \sigma_{M}\sqrt{SC}\left(1 + \rho\right).
\end{equation}

When $\rho$ values are negative, such as when $\textbf{M}$ models a predator-prey system in which $M_{ij}$ and $M_{ji}$ have opposing signs, stability is expected to increase [@Allesina2012]. When elements of $\gamma$ vary independently, the magnitude of $\rho$ is decreased because $\sigma^{2}_{\gamma}$ increases the variance of $M_{ij}$ without affecting the covariance between $M_{ij}$ and $M_{ji}$,

\begin{equation} \label{rho_def}
\rho = \frac{cov(M_{ij}, M_{ji})}{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})}.
\end{equation}

Because $\gamma$ does not affect $cov(M_{ij}, M_{ji})$, the sign of $\rho$ remains unchanged.

**Numerical simulations**. I used numerical simulations to isolate the effect of $\sigma_{\gamma}$ on the stability of $\textbf{M} = \gamma \textbf{A}$ by comparing $\sigma_{\gamma} > 0$ versus $\sigma_{\gamma} = 0$. Values of $\gamma_{ii}$ were sampled such that $E[d]$ and $E[\mu]$ also remained unchanged (in practice, diagonal elements of $\textbf{M}$ were standardised so that mean values were identical before and after adding $\gamma$).

### 2020-03-02 notebook/ms.tex
Rearrange some of the text to improve the flow between analytical and numerical simulation sections of the Results

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (e.g., carrying capacity in ecological communities).
Traditionally, off-diagonal elements are assigned non-zero values with a
probability \(C\), which are sampled from a distribution with variance
\(\sigma^{2}\); diagonal elements are set to
-1\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative
(\(\max\left(\Re(\lambda)\right) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -1\) (the mean value of diagonal elements) and
\(\Im = 0\), with a radius of
\(\sigma\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma^{2}\) increase.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that the expected
response rates (\(\gamma\)) of individual components to perturbations of
the system are identical, but this is highly unlikely in any complex
system. In ecological communities, for example, the rate at which
population density changes following perturbation will depend on the
generation time of organisms, which might vary by orders of magnitude
among species. Species with short generation times will respond quickly
(high \(\gamma\)) to perturbations relative to species with long
generation times (low \(\gamma\)). Similarly, the speed at which
individual banks respond to perturbations in financial networks, or
individuals or institutions respond to perturbations in complex social
networks, is likely to vary. The effect of such variance on stability
has not been investigated in complex systems theory. Intuitively,
variation in \(\gamma\) might be expected to decrease system stability
by introducing a new source of variation into the system and thereby
increasing \(\sigma\). Here I show that, despite higher \(\sigma\),
realistic complex systems (such that \(S\) is high but finite) are
actually more likely to be stable if their individual component response
rates vary. My results are robust across commonly observed network
structures, including
random\textsuperscript{\protect\hyperlink{ref-May1972}{1}},
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}},
cascade food
web\textsuperscript{\protect\hyperlink{ref-Williams2000}{17}} networks.

@@

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma_{\gamma}\)) affect
complex systems. In the absence of such variation
(\(\sigma_{\gamma} = 0\)), \(\gamma\) is set to the identity matrix
(diagonal elements all equal 1) and \(\textbf{M} = \textbf{A}\). Given
this baseline condition, \(\max(\Re(\lambda))\) for
\(\textbf{M} = \textbf{A}\) can be estimated from five
values\textsuperscript{\protect\hyperlink{ref-Tang2014b}{19}}. These
values include (1) system size (\(S\)), (2) mean self-regulation of
components (\(d\); i.e., mean diagonal element of \(\textbf{M}\)), (3)
mean interaction strength between components (\(\mu\)), (4) the standard
deviation between component interaction strengths (\(\sigma\)), and (5)
the correlation of interaction strengths between components, \(M_{ij}\)
and \(M_{ji}\) (\(\rho\)).

When investigating the effect of varying component response rate
\(\sigma_{\gamma}\) on stability by defining
\(\textbf{M} = \gamma\textbf{A}\), \(S\) remains unchanged. Further,
values of \(\gamma_{i}\) were sampled such that \(E[d]\) and \(E[\mu]\)
also remained unchanged (in practice, diagonal elements of
\(\textbf{M}\) were standardised so that mean values were identical
before and after adding \(\gamma\)). What \(\sigma_{\gamma}\) does
change is the variation in component interaction strengths, and
\(\rho\). Introducing variation in \(\gamma\) increases the total
variation in the system, making it more likely that
\(\textbf{M} = \gamma\textbf{A}\) is unstable. Variation of the
off-diagonal elements in \(\textbf{M}\) is described by the joint
variation of two random variables (to simplify the notation,
\(\sigma^{2}_{M}\) and \(\sigma^{2}_{A}\) refer to variances of
off-diagonal elements only),

@@

Note that in my simulations \(E[\gamma_{i}] = 1\) and \(E[A_{ij}] = 0\),
so equation \ref{var_full} can be simplified,

@@

The increase caused by \(\sigma^{2}_\gamma\) can be visualised from the
eigenvalue spectra of \(\textbf{A}\) versus
\(\textbf{M} = \gamma\textbf{A}\). Given \(d = 0\) and \(C = 1\), the
distribution of eigenvalues of \(\textbf{M}\) and \(\textbf{A}\) lie
within a circle of a radius \(\sigma^{2}_{M}\sqrt{S}\) and
\(\sigma^{2}_{A}\sqrt{S}\), respectively.

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.718 and 0.742 for a and b,
respectively.

>>>>

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (\(d\), e.g., carrying capacity in ecological
communities). Traditionally, off-diagonal elements are assigned non-zero
values with a probability \(C\), which are sampled from a distribution
with variance \(\sigma_{M}^{2}\); diagonal elements are set to
\(d = -1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative
(\(\max\left(\Re(\lambda)\right) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -1\) (the mean value of diagonal elements) and
\(\Im = 0\), with a radius of
\(\sigma_{M}\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S_{M}\), \(C_{M}\), and
\(\sigma_{M}^{2}\) increase.

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma_{M}\sqrt{SC} < 1\) assumes that the
expected response rates (\(\gamma\)) of individual components to
perturbations of the system are identical, but this is highly unlikely
in any complex system. In ecological communities, for example, the rate
at which population density changes following perturbation will depend
on the generation time of organisms, which might vary by orders of
magnitude among species. Species with short generation times will
respond quickly (high \(\gamma\)) to perturbations relative to species
with long generation times (low \(\gamma\)). Similarly, the speed at
which individual banks respond to perturbations in financial networks,
or individuals or institutions respond to perturbations in complex
social networks, is likely to vary. The effect of such variance on
stability has not been investigated in complex systems theory.
Intuitively, variation in \(\gamma\) (\(\sigma^{2}_{\gamma}\)) might be
expected to decrease system stability by introducing a new source of
variation into the system and thereby increasing \(\sigma_{M}\). Here I
show that, despite higher \(\sigma_{M}\), realistic complex systems
(such that \(S\) is high but finite) are actually more likely to be
stable if their individual component response rates vary. My results are
robust across commonly observed network structures, including
random\textsuperscript{\protect\hyperlink{ref-May1972}{1}},
small-world\textsuperscript{\protect\hyperlink{ref-Watts1998}{15}},
scale-free\textsuperscript{\protect\hyperlink{ref-Albert2002}{16}},
cascade food
web\textsuperscript{\protect\hyperlink{ref-Williams2000}{17}} networks.

@@

Equation \ref{defM} thereby serves as a null model to investigate how
variation in component response rate (\(\sigma_{\gamma}\)) affects
complex systems. In the absence of such variation
(\(\sigma_{\gamma} = 0\)), \(\gamma\) is set to the identity matrix
(diagonal elements all equal 1) and \(\textbf{M} = \textbf{A}\).

\textbf{Analytical considerations}. The value of \(\max(\Re(\lambda))\),
and therefore system stability, can be estimated from five properties of
\(\textbf{M}\)\textsuperscript{\protect\hyperlink{ref-Tang2014b}{19}}.
These properties include (1) system size (\(S\)), (2) mean
self-regulation of components (\(d\)), (3) mean interaction strength
between components (\(\mu\)), (4) the standard deviation between
component interaction strengths (\(\sigma_{M}\)), and (5) the
correlation of interaction strengths between components, \(M_{ij}\) and
\(M_{ji}\) (\(\rho\)). Positive \(\sigma_{\gamma}\) does not change
\(S\), nor does it necessarily change \(E[d]\) or \(E[\mu]\). What
\(\sigma_{\gamma}\) does change is the total variation in component
interaction strengths (\(\sigma_{M}\)), and \(\rho\). Introducing
variation in \(\gamma\) increases the total variation in the system.
Variation of the off-diagonal elements in \(\textbf{M}\) is described by
the joint variation of two random variables,

@@

Given \(E[\gamma_{i}] = 1\) and \(E[A_{ij}] = 0\), equation
\ref{var_full} can be simplified,

@@

The increase caused by \(\sigma^{2}_\gamma\) can be visualised from the
eigenvalue spectra of \(\textbf{A}\) versus
\(\textbf{M} = \gamma\textbf{A}\) (Figure 1). Given \(d = 0\) and
\(C = 1\), the distribution of eigenvalues of \(\textbf{A}\) and
\(\textbf{M}\) lie within a circle of a radius
\(\sigma^{2}_{A}\sqrt{S}\) and \(\sigma^{2}_{M}\sqrt{S}\), respectively
(Figure 1a vs.~1b). When components are self-regulating (e.g.,
\(d = -1\)), positive \(\sigma^{2}_\gamma\) changes the distribution of
eigenvalues, potentially affecting stability (Figure 1c vs.~1d).

Given \(\sigma^{2}_\gamma = 0\), \(\max(\Re(\lambda))\) decrease
linearly with \(\rho\) such that,

\begin{equation} \label{rho_stab}
\max(\Re(\lambda)) \approx \sigma_{M}\sqrt{SC}\left(1 + \rho\right).
\end{equation}

When \(\rho\) values are negative, such as when \(\textbf{M}\) models a
predator-prey system in which \(M_{ij}\) and \(M_{ji}\) have opposing
signs, stability is expected to
increase\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. When
elements of \(\gamma\) vary independently, the magnitude of \(\rho\) is
decreased because \(\sigma^{2}_{\gamma}\) increases the variance of
\(M_{ij}\) without affecting the covariance between \(M_{ij}\) and
\(M_{ji}\),

\begin{equation} \label{rho_def}
\rho = \frac{cov(M_{ij}, M_{ji})}{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})}.
\end{equation}

Because \(\gamma\) does not affect \(cov(M_{ij}, M_{ji})\), the sign of
\(\rho\) remains unchanged.

\textbf{Numerical simulations}. I used numerical simulations to isolate
the effect of \(\sigma_{\gamma}\) on the stability of
\(\textbf{M} = \gamma \textbf{A}\) by comparing \(\sigma_{\gamma} > 0\)
versus \(\sigma_{\gamma} = 0\). Values of \(\gamma_{ii}\) were sampled
such that \(E[d]\) and \(E[\mu]\) also remained unchanged (in practice,
diagonal elements of \(\textbf{M}\) were standardised so that mean
values were identical before and after adding \(\gamma\)).

@@

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 0.05\), and \(\sigma = 0.4\).
\textbf{a.} Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but also creates a cluster of
eigenvalues toward the distribution's centre (-1, 0). Blue elipses in
both panels show the circle centred on the distribution in panel a.
Proportions of \(\Re(\lambda) < 0\) are 0.727 and 0.733 for a and b,
respectively.
